IML Term project paper

### 2022-12-11 paper.rmd
Update paper.rmd
Pieniä korjauksia, lisätty dummy-malli, mallin valinta ja multiclass-classifier-osio

@@

ogistic regression models can be used for binomial variables as well as for multinomial variables under specific assumptions. The idea of logistic regression in a binary case is that our variable of interest $Y$ is binomially distributed with parameter $q\in [0,1]$ representing probability that $Y$ receieves one of the values (this is somewhat irrelevant as the corresponding parameter for the other value would simply be $1-q$). Now, with explanatory variables $X$ we assume that these $X$ influence the parameter $p$ for each observation, meaning $p(Y|X)=q(X)^{Y}(1-q(X))^{1-Y}$ is a binomial probability with parameter $q(X)$ being of form $\sigma(\sum \lambda_i X_i)$ for values of individual explanatory variables $X_i$ and coefficients $\lambda_i$ and link function $\sigma: \mathbb{R}\to [0,1]$. Thus we can estimate a logistic regression by finding parameters $\lambda_i$ that maximize the likelihood function $\prod p(Y_i|X)$.

@@

However, possible shortcomings of the logistic regression model include assumptions of linearity in relation to the explanatory variables and the parameter (which we could attempt to correct but with such a collection of explanatory variables, this at least can't be applied immediately), assumptions of for link function that are required to define coefficients, and especially with high-dimensional models, multicollinearity can become a notable problem.

In our model, we used specifically the logistic link function, implying $\sigma(x)=\frac{e^{x}}{e^{x}+1}$, although we acknowledge that other choices such as the probit link could have also been viable options. In order to curb issues related to multicollinearity, we first limited ourselves to the dataset with observations measured at multiples only observed at 16.8 meters. After this, we reduced the set of explanatory variables further via variance inflation factors. Variance inflation factors (VIFs) can be calculated for each explanatory variable in a model through assessing linear models for each of the explanatory variables. The VIF for explanatory variable $X_i$ is defined as $\frac{1}{1-R_i^2}$ where $R_i^2$ is the amount of variance of $X_i$ explained through a linear model by the other explanatory variables. Thus, these values express how much the explanatory variable in question can be described by a linear combination of the other explanatory variables. Thus, particularly high VIFs imply multicollinearity being present in the model.

@@

In terms of our modelling, we decided to use lasso instead of a comparable method called ridge regression, which uses penalty term $\sum_{i=1}^n \gamma \lambda_i^2$. The reason for this is that we preferred to make a simpler model with fewer explanatory variables rather than ridge regression, which tends to not reduce the dimension of the model but rather simply give similar coefficients to all correlated explanatory variables. We attempted to assess the value of $\gamma$ which would minimize loo cross-validated misclassification rate with the function cv.glmnet. After defining this value, we calculated cross-validated accuracy (which should be equivalent to what was calculated by cv.glmnet) as well as the cross-validated perplexity.

@@

For our model, we chose to run a 10-fold cross validation on a set of tuning parameters $C$ to assess which kernel performed best and chose to model the value based on that. In these preliminary assessments, we assessed the radial and linear kernels had the best performance of the ones available in the R package e1071. For these kernels, we ran loo-cross validation to assess the value of tuning parameter $C$ and calculated for said tuning parameter loo-cross-validated accuracy and perplexity. Eventually, the radial kernel had slightly better accuracy but the linear kernel had lower perplexity and as a result, we chose to use the linear kernel in our model

@@

@@

>>>>

### Dummy model

As our first model, we simply attempted to predict the values through a dummy model, which predicts all values into the more common of the two classes. However, as half of the observations are nonevent days and half of the observations are event days, accuracy would only be 50% no matter which the model would choose. As a result, the dummy model is not sufficient to use in this problem

@@

Logistic regression models can be used for binomial variables as well as for multinomial variables under specific assumptions. The idea of logistic regression in a binary case is that our variable of interest $Y$ is binomially distributed with parameter $q\in [0,1]$ representing probability that $Y$ receieves one of the values. Now, with explanatory variables $X$ we assume that these $X$ influence the parameter $p$ for each observation, meaning $p(Y|X)=q(X)^{Y}(1-q(X))^{1-Y}$ is a binomial probability with parameter $q(X)$ being of form $\sigma(\sum \lambda_i X_i)$ for values of individual explanatory variables $X_i$ and coefficients $\lambda_i$ and link function $\sigma: \mathbb{R}\to [0,1]$. Thus we can estimate a logistic regression by finding parameters $\lambda_i$ that maximize the likelihood function $\prod p(Y_i|X)$.

@@

However, possible shortcomings of the logistic regression model include assumptions of linearity in relation to the explanatory variables and the parameter (which we could attempt to correct but with such a large collection of explanatory variables, this at least can't be applied immediately), assumptions of for example link function that are required to define coefficients, and especially with high-dimensional models, multicollinearity can become a notable problem.

In our model, we used specifically the logistic link function, implying $\sigma(x)=\frac{e^{x}}{e^{x}+1}$, although we acknowledge that other choices such as the probit link could have also been viable options. The estimation of the model was done with base r function glm. In order to curb issues related to multicollinearity, we first limited ourselves to the dataset with observations measured at multiples only observed at 16.8 meters. After this, we reduced the set of explanatory variables further via variance inflation factors.

Variance inflation factors (VIFs) can be calculated for each explanatory variable in a model through assessing linear models for each of the explanatory variables. The VIF for explanatory variable $X_i$ is defined as $\frac{1}{1-R_i^2}$ where $R_i^2$ is the amount of variance of $X_i$ explained through a linear model by the other explanatory variables. Thus, these values express how much the explanatory variable in question can be described by a linear combination of the other explanatory variables. Thus, particularly high VIFs imply multicollinearity being present in the model. We calculated VIFs for our model with function vif from R package car.

@@

In terms of our modelling, we decided to use lasso instead of a comparable method called ridge regression, which uses penalty term $\sum_{i=1}^n \gamma \lambda_i^2$. The reason for this is that we preferred to make a simpler model with fewer explanatory variables rather than ridge regression, which tends to not reduce the dimension of the model but rather simply give similar coefficients to all correlated explanatory variables. In order to estimate the model, we used R functions glmnet and cv.glmnet from package glmnet. We attempted to assess the value of $\gamma$ which would minimize loo cross-validated misclassification rate with the function cv.glmnet. After defining this value, we calculated cross-validated accuracy (which should be equivalent to what was calculated by cv.glmnet) as well as the cross-validated perplexity for the model estimated by glmnet.

@@

For our model, we used functions tune and svm from R package e1071. We chose to run a 10-fold cross validation on a set of tuning parameters $C$ to assess which kernel performed best and chose to model the value based on that. In these preliminary assessments, we assessed the radial and linear kernels had the best performance of the ones available in package e1071. For these kernels, we ran loo-cross validation to assess the value of tuning parameter $C$ and calculated for said tuning parameter loo-cross-validated accuracy and perplexity. Eventually, the radial kernel had slightly better accuracy but the linear kernel had lower perplexity and as a result, we chose to use the linear kernel in our model.

@@

With our observed best-performing models being the logistic regression model, the Lasso regression model and the linear discriminant analysis. Out of these models, we eventually chose to use the logistic regresison model. The Lasso model was our best-performing model, but we eventually chose not to use it for practical reasons related to computational heaviness in calculating the multi-class model. Mainly, the amount of information we could gain through calculations in R without R crashing was relatively limited, to the degree that we couldn't always even calculate the cross-validated best choice for the tuning parameter.

We eventually also chose to not use the LDA model after assessing the assumptions related to the model. LDA models have been found to perform best with a limited amount of variables and we still question whether we should attempt to reduce the amount of variables from the model further from the current amount of 38 from the model. On top of this, considering the assumption of equal variances, the boxplots we drew earlier in the report imply that class "nonevent" might have larger variance for some variables than class "event".

Thus we were left with the logistic regression model, which was also lower-dimensional than many of the other models with only 10 explanatory variables. However, we could question in relation to this model whether there is sufficient basis to for example remove certain variables, which might describe relevant information, as well as for example whether including an intercept in the model would be meaningful in terms of interpretation. Currently, several of these decisions are based on likelihood-ratio tests performed within the R function glm, which was pursued in an attempt to create a lower-dimensional model. Also, comparing the logistic regression model to for example the Lasso, which does variable selection on a model with a somewhat comparable minimzable loss function.

## Multiclass-classifier
we will extend it into a multinomial logistic regression model in order to assess the multi-class accuracy for our model. The basic idea of multinomial logistic regression here is that out of $K$ classes we choose a baseline value of our variable of interest, and compare other values to it, leading to $p(Y=k|X=x)=\frac{e^{\beta_{k0}+\sum_{j=1}^m\beta_{kj}x_{ij}}}{1+\sum_{k=1}^{K-1} e^{\beta_{k0}+\sum_{j=1}^m\beta_{kj}x_{ij}}}$ where our baseline class is class $K$. Alternatively, we can use equivalent so-called softmax coding where we  treat all classes equivalently and thus $p(Y=k|X=x)=\frac{e^{\beta_{k0}+\sum_{j=1}^m\beta_{kj}x_{ij}}}{\sum_{k=1}^K-1 e^{\beta_{k0}+\sum_{j=1}^m\beta_{kj}x_{ij}}}$. The interpretation of logistic regression coefficients comes now from the following: $\log(\frac{p(\hat{y}_i=k_a)}{p(\hat{y_i}=k_b)})=(\lambda_{a0}-\lambda_{b0})+\sum_{j=1}^n(\lambda_{aj}-\lambda_{bj})x_{ij}$. Thus the interpretation of these coefficients is in relation to the difference of the coefficients for different classes and describe the ratio between the probabilities that our variable of interest receives a value in one of these two classes.

Similarly as in logistic regression, we attempt to define coefficients $\lambda_j$ in a way that maximizes likelihood.

In our model, we used the same dataset as in our binary logistic regression model, in order to avoid issues of multicollinearity. In order to define this model, we used function multinom from R package nnet with softmax codin. This creates estimated probabilities $\hat{p}(\hat{y}_i=y)$ for all four classes $y$, and through that we could calculate loo-cross-validated values for accuracy as well as perplexity.

Eventually, we chose to emphasize class2 accuracy, and as a result our final predictions are based on estimating models for both class2 and class4. We first used our model for class2 to predict whether a value would be an event or not and for values the model assessing class2 predicted to be event-days, we referred to our class4 model to estimate which of the event classes had the highest probability.

@@

We notice in terms of multiclass accuracy that while the model predicts in particular nonevents rather well, the loo-cross-validated predictions are less successful for calculating distinct event types. For example, less than 10% of days which belonged to class Ia were predicted into this class. As a result, further changes should be made to the model in the future to make these predictions more viable.
